---
title: "DATA 606 Assignment 2"
author:
- "Michael Ellsworth"
- 'UCID: 30101253'
date: "Februrary 9th, 2020"
output:
  html_document:
    df_print: paged
---

# Explore the dataset: Arthritis
```{r Load Packages, include = FALSE}
library(vcd)
library(dplyr)
library(fmsb)
data(Arthritis)
head(Arthritis)
```

*Now in the "Improved" column, there are three response values: "None", "Some" and "Marked". Next we construct another column indicating whether or not the arthritis gets improved.*

```{r Create new Data Frame with Indicator column}
newDATA <- Arthritis %>%
  mutate(Indicator = ifelse(Improved == "None", "No", "Yes"))
head(newDATA)
```

# Question 1

*Please create a table and use three different tests (risk difference, risk ratio, odds ratio) to test whether the treatment can help improve arthritis.*

```{r Question 1}
# Load in the Publish package
library(Publish)

# Create a table with Treatment and Indicator
newDATA_table <- table(newDATA$Treatment, newDATA$Indicator)

# Use the table2x2 function to output tests
table2x2(newDATA_table)
```

From the `table2x2` function, we can output all of the desired tests.

The estimated risk difference from the `table2x2` output is 35.7% with a 95% confidence interval ranging from 15.5% to 55.7%.

The estimate risk ratio from the `table2x2` output is 2.127 with a 95% confidence interval ranging from 1.297 and 3.489.

The estimated odds ratio from the `table2x2` output is 4.462 with a 95% confidence interval ranging from 1.785 and 11.154.

# Question 2

*Now we doubt if the treatment is conditional independent with the improvement. Use odds ratio test to test whether the treatment can improve arthritis conditional on the sex.*

```{r Question 2}
# Create a new table for Treatment and Indicator where sex is equal to male
male <- newDATA %>% filter(Sex == "Male")
male_table <- table(male$Treatment, male$Indicator)

# Create a new table for Treatment and Indicator where sex is equal to female
female <- newDATA %>% filter(Sex == "Female")
female_table <- table(female$Treatment, female$Indicator)

# Compute the odds ratio for both the male and female tables
male_odds <- oddsratio(male_table, conf.level = 0.95, p.calc.by.independence = TRUE)
female_odds <- oddsratio(female_table, conf.level = 0.95, p.calc.by.independence = TRUE)
male_odds
female_odds
```

Both odds ratio estimates are significant (p-value is < 0.05). Improvement is not conditional on sex.

*To explore the influence of age to the improvement, we do the following things first.*

```{r Create new Data Frame with Age Level column}
# We order the dataset as per the age
DATA<-Arthritis[order(Arthritis$Age),]
# We group the patients as per their ages (20-39, 40-59, 60-79)
l1=sum(as.numeric(DATA$Age<=39))
l2=sum(as.numeric(DATA$Age<=59))-l1
l3=dim(DATA)[1]-l1-l2
Age_level<-c(rep('20-39', l1), rep('40-59', l2), rep('60-79', l3))
myDATA<-cbind(Arthritis, Age_level)
head(myDATA)
```

# Question 3

*Apply Pearson Chi-square test to test whether or not Age_level affects the improvement (based on myDATA)*

```{r Question 3}
age_improvement_tab <- table(myDATA$Improved, myDATA$Age_level)
chisq.test(age_improvement_tab)
```

Based on a p-value > 0.05, we cannot say that age level affects the improvement.

# Question 4

*Following Q3, can you compute the Pearson standardized residuals to see which cell deviates most from the independence assumption*

```{r}
library(questionr)
chisq.residuals(age_improvement_tab, std = TRUE)
```

Based on the standardized residuals, the 60-79 age level deviates most from the independence assumption.

# Question 5

*Following Q3, as both row and column variables are ordinal, please also perform Mantel-Haenszel test (choose the scores by yourself). Compare your result with the Chi-square test*

```{r}
pears.cor=function(table, rscore, cscore)
{ 
	dim=dim(table) 
	rbar=sum(margin.table(table,1)*rscore)/sum(table) 
	rdif=rscore-rbar 
	cbar=sum(margin.table(table,2)*cscore)/sum(table) 
	cdif=cscore-cbar 
	ssr=sum(margin.table(table,1)*(rdif^2)) 
	ssc=sum(margin.table(table,2)*(cdif^2)) 
	ssrc=sum(t(table*rdif)*cdif) 
	pcor=ssrc/(sqrt(ssr*ssc)) 
	pcor 
	M2=(sum(table)-1)*pcor^2
	M2
	result=c(pcor, M2, (1-pchisq(M2,1)))
	result=as.table(result)
	names(result)=c('Pearson correlation','MH statistic', 'P-Value')
	result
}
pears.cor(age_improvement_tab, c(1, 2, 3), c(30, 50, 70))
```

Since the p-value is > 0.05, we cannot say that age level affects the improvement. This is similar to the Chi-square test.

# Question 6

*Did Chi-square test give you an accurate test result? If not, use Fisher's exact test to justify the conclusion from Chi-square test.*

```{r}
fisher.test(age_improvement_tab, alternative = "two.sided")
```

The fisher test confirms the result from the Chi-square test. Age level does not affect improvement.
