---
title: "Michael Ellsworth - Assignment 1"
subtitle: "30101253"
output:
  html_document:
    df_print: paged
---
# Further explore the dataset "api"

The dataset "api" in package "survey" contains the academic performance index for year 1999 (api99) and 2000 (api00). To import the data, you need to attach the package "survey".

```{r include = FALSE}
library(survey)
library(sampling)
library(dplyr)
library(ggplot2)
data(api)
API <- apipop[-c(seq(1,4)),] # remove the first four units

# Sort the dataset according to dnum
API <- API[order(API$dnum), ]

# Construct clusters based on dnum, each contains 619 units (equal size)
L <- matrix(c(rep("L1",619),
              rep("L2",619),
              rep("L3",619),
              rep("L4",619),
              rep("L5",619),
              rep("L6",619),
              rep("L7",619),
              rep("L8",619),
              rep("L9",619),
              rep("L10",619)),
            length(API$dnum),1)
MYDATA <- cbind(L, API)

# The name for clusters is "cluster_name"
names(MYDATA)[1] <- c("cluster_name")
MYDATA <- MYDATA %>% select(cluster_name, api00, grad.sch, api99)
head(MYDATA)
```

Now treat the new constructed dataset "MYDATA" as the "population" source. **The variable** we are interested is **"api00"**.

*Goal*: **estimate the population average as well as the standard deviation of the estimate**.

## Population level

### Question 1
Calculate the population mean

```{r Quesetion 1 Population Mean}
pop_mean <- mean(MYDATA$api00)
pop_mean
```

## Equal-size cluster sampling

### Question 2
Apply one-stage cluster sampling to estimate the population mean (**n=3**), calculate its estimated standard deviation (refer to the formula in lecture notes), **using L1$\sim$L10 as clusters**.

```{r Question 2 Population Estimates}
# Define number of clusters sampled, number of clusters total and size of clusters
n_cluster <- 3
N_cluster <- 10
M_cluster <- nrow(MYDATA)/N_cluster

# Create a vector of cluster names to sample from
clusters <- paste0("L", seq(1, 10))

# Sample from the cluster name vectors n times
cluster_sample <- sample(clusters, n_cluster)

# Filter the MYDATA table to include only the clusters from cluster_sample
MYDATA_cluster_sample <- MYDATA %>%
  filter(cluster_name %in% cluster_sample)

# Calculate the totals from each of the clusters
results <- MYDATA_cluster_sample %>%
  group_by(cluster_name) %>%
  summarise(totals = sum(api00), count = n())

# Estimate the population mean
mean(MYDATA_cluster_sample$api00)

# Estimate the standard deviation
((1 - n_cluster / N_cluster) * var(results$totals) / (n_cluster * M_cluster**2))**0.5
```

### Question 3
Please repeat Q2 10 times which gives you at least 10 population mean estimates (**e.g. using "for" loop**), compute the average of these 10 estimates as well as their standard deviation. 
```{r Question 3 Population Estimates Mean Cluster}
# Initialize a vector of population mean estimates
pop_mean_estimates_cluster <- c()

# Create a for loop to append pop_mean_estimates
for(i in 1:10){
  # Sample from the cluster name vectors n times
  cluster_sample <- sample(clusters, n_cluster)
  
  # Filter the MYDATA table to include only the clusters from cluster_sample
  MYDATA_cluster_sample <- MYDATA %>%
    filter(cluster_name %in% cluster_sample)
  
  # Estimate the population mean
  pop_mean_estimates_cluster[i] <- mean(MYDATA_cluster_sample$api00)
}
# Ten estimates of the population mean
pop_mean_estimates_cluster

# Calculate the mean of the pop_mean_estimates_cluster
mean_pop_estimates_cluster <- mean(pop_mean_estimates_cluster)
mean_pop_estimates_cluster

# Calculate the standard deviation of the pop_mean_estimates_cluster
sd_pop_estimates_cluster <- sd(pop_mean_estimates_cluster)
sd_pop_estimates_cluster
```

### Question 4
Take an SRS (simple random sample) of the same sample size (3*619) to estimate the population mean, calculate its estimated standard deviation (refer to the formula in lecture notes).
```{r Question 4 Simple Random Sample}
# Define the secondary sampling unit sample size
n_SRS <- 619 * 3
N_SRS <- 619 * 10

# Filter the MYDATA table to include only the clusters from cluster_sample
MYDATA_SRS <- sample_n(MYDATA, n_SRS)

# Estimate the population mean
mean(MYDATA_SRS$api00)

# Estimate the standard deviation
sd(MYDATA_SRS$api00)
```

### Question 5
Repeat Q4 for at least 10 times which gives you at least 10 population mean estimates (**e.g. using "for" loop**), compute the average of these 10 estimates as well as their standard deviation.
```{r Question 5 Population Estimates Mean SRS}
# Initialize a vector of population mean estimates
pop_mean_estimates_SRS <- c()

# Create a for loop to append pop_mean_estimates
for(i in 1:10){
  # Filter the MYDATA table to include only the clusters from cluster_sample
  MYDATA_SRS <- sample_n(MYDATA, n_SRS)
  
  # Estimate the population mean
  pop_mean_estimates_SRS[i] <- mean(MYDATA_SRS$api00)
}
# Ten population estimates
pop_mean_estimates_SRS

# Calculate the mean of the pop_mean_estimates_SRS
mean_pop_estimates_SRS <- mean(pop_mean_estimates_SRS)
mean_pop_estimates_SRS

# Calculate the standard deviation of the pop_mean_estimates_SRS
sd_pop_estimates_SRS <- sd(pop_mean_estimates_SRS)
sd_pop_estimates_SRS
```

### Question 6
Compare your estimates (cluster-sampling and SRS) with the population mean, **use ANOVA table (calculate SSB and SSW)** to explain why your cluster-sampling-based estimate is good or bad.

```{r Question 6 SSB and SSW}
# Find population means in each cluster
SSB_data <- MYDATA %>%
  group_by(cluster_name) %>%
  summarise(mean = mean(api00)) %>%
  mutate(square_diff = (mean - pop_mean)^2)

# Sum the squares (SSB) between psus
SSB <- sum(SSB_data$square_diff) * 619
SSB
MSB <- SSB / (N_cluster - 1)

# Create a data set with cluster means and differences squared
SSW_data <- MYDATA %>%
  group_by(cluster_name) %>%
  mutate(cluster_mean = mean(api00)) %>%
  ungroup() %>%
  mutate(square_diff = (api00 - cluster_mean)^2)

# Sum the squares (SSW) within psus
SSW <- sum(SSW_data$square_diff)
SSW
MSW <- SSW / (N_cluster * (M_cluster - 1))

# Total
SSTO <- SSW + SSB
S_squared <- SSTO / (N_cluster * M_cluster - 1)

# Anova table for cluster sampling
summary(aov(api00 ~ cluster_name, data = MYDATA))

# Is MSB greater than S^2?
MSB > S_squared
```

Since MSB is greater than $S^2$, cluster sampling is more volatile than simple random sample.


## Ratio estimation
### Question 7
Use variable **grad.sch** as auxiliary variable, apply the ratio estimation to estimate the population mean (take SRS, get B_hat and then estimate).
```{r Question 7 Estimate Population Mean RE using grad.sch}
# Take a simple random sample of MYDATA using sample size of n = 3 * 619
MYDATA_SRS_7 <- sample_n(MYDATA, n_SRS)

# Calculate means of the sample of y (api00) and x (grad.sch)
sample_mean_api00 <- mean(MYDATA_SRS_7$api00)
sample_mean_grad.sch <- mean(MYDATA_SRS_7$grad.sch)

# Calculate the ratio estimate (B_hat)
B_hat <- sample_mean_api00 / sample_mean_grad.sch

# Calculate the ratio estimate of the population mean of y (api00) using the known population mean of x (grad.sch)
mean_api00_ratio_estimate <- B_hat * mean(MYDATA$grad.sch)
mean_api00_ratio_estimate
```


### Question 8
Repeat the above procedure for at least 10 times (**e.g. using "for" loop**), get at least 10 estimates, compute their average as well as standard deviation.
```{r Question 8 Mean and SD of Estimates of Population Means}
mean_api00_ratio_estimate_vector_8 <- c()

for(i in 1:10){
  # Take a simple random sample of MYDATA using sample size of n = 3 * 619
  MYDATA_SRS_8 <- sample_n(MYDATA, n_SRS)
  
  # Calculate means of the sample of y (api00) and x (grad.sch)
  sample_mean_api00 <- mean(MYDATA_SRS_8$api00)
  sample_mean_grad.sch <- mean(MYDATA_SRS_8$grad.sch)
  
  # Calculate the ratio estimate (B_hat)
  B_hat <- sample_mean_api00 / sample_mean_grad.sch
  
  # Calculate the ratio estimates of the population mean of y (api00) using the known population mean of x (grad.sch)
  mean_api00_ratio_estimate_vector_8[i] <- B_hat * mean(MYDATA$grad.sch)
}
# Ten estimates of the population mean
mean_api00_ratio_estimate_vector_8

# Mean of the ten estimates of the population mean
mean(mean_api00_ratio_estimate_vector_8)

# Standard deviation of the ten estimates of the population mean
sd(mean_api00_ratio_estimate_vector_8)
```

### Question 9
Use variable **api99** as auxiliary variable, apply the ratio estimation to estimate the population mean (take SRS, get B_hat and then estimate).
```{r Question 9 Estimate Population Mean RE using api99}
# Take a simple random sample of MYDATA using sample size of n = 3 * 619
MYDATA_SRS_9 <- sample_n(MYDATA, n_SRS)

# Calculate means of the sample of y (api00) and x (api99)
sample_mean_api00 <- mean(MYDATA_SRS_9$api00)
sample_mean_api99 <- mean(MYDATA_SRS_9$api99)

# Calculate the ratio estimate (B_hat)
B_hat <- sample_mean_api00 / sample_mean_api99

# Calculate the ratio estimate of the population mean of y (api00) using the known population mean of x (grad.sch)
mean_api00_ratio_estimate <- B_hat * mean(MYDATA$api99)
mean_api00_ratio_estimate
```

### Question 10
Repeat the above procedure for at least 10 times (**e.g. using "for" loop**), get at least 10 estimates, compute their average as well as standard deviation.
```{r Question 10 Mean and SD of Estimates of Population Means}
mean_api00_ratio_estimate_vector_10 <- c()

for(i in 1:10){
  # Take a simple random sample of MYDATA using sample size of n = 3 * 619
  MYDATA_SRS_10 <- sample_n(MYDATA, n_SRS)
  
  # Calculate means of the sample of y (api00) and x (grad.sch)
  sample_mean_api00 <- mean(MYDATA_SRS_10$api00)
  sample_mean_api99 <- mean(MYDATA_SRS_10$api99)
  
  # Calculate the ratio estimate (B_hat)
  B_hat <- sample_mean_api00 / sample_mean_api99
  
  # Calculate the ratio estimates of the population mean of y (api00) using the known population mean of x (api99)
  mean_api00_ratio_estimate_vector_10[i] <- B_hat * mean(MYDATA$api99)
}
# Ten estimates of the population mean
mean_api00_ratio_estimate_vector_10

# Mean of the ten estimates of the population mean
mean(mean_api00_ratio_estimate_vector_10)

# Standard deviation of the ten estimates of the population mean
sd(mean_api00_ratio_estimate_vector_10)
```

### Question 11
Compare the estimation results above, which auxiliary variable gives better estimation? Explain why (do significance test, calculate correlation coefficient).


The ratio estimate of the population mean of api00 is better estimated using the auxiliary variable api00 than grad.sch. This is evident from the standard deviation of each of the population mean estimates calculated in Question 8 and Question 10. The standard deviation of the population mean estimates using grad.sch is ~`{r} sd(mean_api00_ratio_estimate_vector_8)` which is much greater than the standard deviation of the population mean estimates using api99, which is ~`{r} sd(mean_api00_ratio_estimate_vector_10)`.

This can also be explained using the correlation coefficient.
```{r Comparison of ratio estimations}
# Calculate the correlation coefficient of api00 ~ grad.sch
cor(x = MYDATA$grad.sch, y = MYDATA$api00)

# Calculate the correlation coefficient of api00 ~ api.99
cor(x = MYDATA$api99, y = MYDATA$api00)
```

Since the correlation coefficient of `api00 ~ api.99` is closer to 1 than `api00 ~ grad.sch`, the relationship is more linear and therefore a better estimator.

### Question 12
Please use function "scatter.smooth" to visualize your explanations in Q11.
**hint: help(scatter.smooth)**


The differences in the correlation coefficient between `api00 ~ grad.sch` and `api00 ~ api.99` can also be expressed visually using scatter plots.
```{r}
# Plot the relationship between grad.sch and api00 to visualize how linear the relationship is
MYDATA %>% ggplot(aes(x = grad.sch, y = api00)) +
  geom_point() + 
  geom_smooth()

# Plot the relationship between api99 and api00 to visualize how linear the relationship is
MYDATA %>% ggplot(aes(x = api99, y = api00)) +
  geom_point() + 
  geom_smooth()
```

As shown by the above plots, the relationship between `api00 ~ api99` is very linear whereas the relationship between `api00 ~ grad.sch` is less linear.