---
title: "DATA 602 - Introduction to Statistical Modeling from Bivariate Data (October 8)"
output: html_notebook
---
&copy; Jim Stallard 2019

```{r include=FALSE}
# Load packages
library(dplyr)
library(ggplot2)
library(mosaic)
library(mosaicData)
library(ISLR)
library(mdsr)
```

# What if the Statistical Result is Significant? Inferences Concerning the Values of $A$ and $B$ Terms in the Linear Model


### 2A. Inferences About $A$

Due to the approximate Normal distributions of the statistics $A$ and $B$ *AND* that we use $S_{e}$ to estimate the common standard deviation in the response variable $Y$, $a$ and $b$ can each be transformed in the following manner:

$$
\frac{a - A}{S_{a}} \hspace{0.5in} {\rm and} \hspace{0.5in} \frac{b - B}{S_{b}}
$$

*Each* of these tranformations is a pitoval quantity and *each* is transformation can be approximated by the Student's $t$ distribuiton with $df = n - 2$. That is, 

$$
T_{df=(n-2)} = \frac{a - A}{S_{a}} \hspace{0.5in} {\rm and} \hspace{0.5in} T_{df = (n - 2)} = \frac{b - B}{S_{b}}
$$

where the **empirical** standard devaitions of $a$ and $b$, $S_{a}$ and $S_{b}$ *can be* computed with the following formulae:


$$
S_{a} = S_{e}\sqrt{\frac{1}{n} + \frac{\overline{X}^{2}}{\sum_{i = 1}^{n}X_{i}^{2} - n\overline{X}^{2}}}
\hspace{0.5in} {\rm and} \hspace{0.5in} S_{b} = \frac{S_{e}}{\sqrt{{\sum_{i = 1}^{n}X_{i}^{2} - n\overline{X}^{2}}}}
$$
 
We will *not be* demontrating how these can be computed, as R will compute these statistics within the `lm(y~x)` assignment. 
</br>
</br>


We can use the above transformation of $a$ to the $T_{n-2}$ to create a (i) confidence interval for $A$ and (ii) conduct a hypothesis test about the value of $A$. 

#### Confidence Interval Estimation of $A$

A 95% confidence interval for the $Y$-intercept term of the linear model, $A$, is 
$$
a \pm \left(t_{0.975,n-2}*S_{a}\right)
$$

**Demonstration 1: ** Compute a 95% confidence interval for the $Y$-intercept term of the model 
$$
AverageSATMathScore_{i} = A + B*AverageSalary_{i} + e_{i}
$$
</br>
</br>
**Answer: ** From what we know, $a=456.83624$. To obtain the value of $S_{a}$, 
```{r}
SAT_2010 = SAT_2010 %>%
  mutate(sat_percategory = cut(sat_pct,
                               breaks = c(0, 33, 67, 100),
                               labels = c("low", "med", "high")))
predictmathSAT <- lm(math~salary, filter(SAT_2010, sat_percategory == "high")) 
coefficients(summary(predictmathSAT))
```
the value in the "Std.Error" column associated with the "(Intercept)" is the value of $S_{a} = 27.70037$.
</br>
</br>
From R we can obtain the 97.5th percentile of the Student's $t$ distribution with $df = 14 - 2 = 12$ degrees of freedom:
```{r}
qt(0.975,14-2)
```
and $t_{0.975,14 - 2} = 2.178813$. The 95% confidence interval for $A$ is then
$$
a \pm \left(t_{0.975,n-2}*S_{a}\right) \longrightarrow 456.83624 \pm (2.178813*27.70037) \longrightarrow 456.83624 \pm 60.3544 \longrightarrow 396.482 \leq A \leq 517.191
$$
How do we interpret the meaning of this interval?
</br>
</br>
**Interpretation:**   
<div style="margin-bottom:200px;">
</div>

This interval can also be computed in R with the `confint(modelname)`:
```{r}
confint(predictmathSAT, level=0.95) #option level=0.95 is the default
```
From this output,
$$
396.4823 \leq A \leq 517.1902
$$

#### Hypothesis Testing about the value of $A$

There are three different statistical hypotheses that can be carried out on the value of the $Y$-intercept term:
$$
\begin{array}{lccc}
{\rm Right-Tailed}                   & {\rm Left-Tailed}       & {\rm Two-Tailed}  \\
{\rm H}_{0}: A \leq A^{*}  & {\rm H}_{0}: A \geq A^{*}  & {\rm H}_{0}: A = A^{*} \\
{\rm H}_{A}: A > A^{*}     & {\rm H}_{A}: A < A^{*}     & {\rm H}_{A}: A \ne A^{*}
\end{array}
$$

The test statistic is computed with $T_{Obs}$
$$
T_{obs} = \frac{a - A^{*}}{S_{a}} \sim T_{n -2}
$$

**Demonstration 2A:** Consider the statsitical hypotheses
$$
{\rm H}_{0}: A = 0 \:\: \text{(the mean value of Y is equal to 0 when x = 0)} \\
{\rm H}_{A}: A > 0 \:\: \text{(the mean value of Y is greater than 0 when x = 0)} \\
$$
To test this, the $T_{Os}$ is computed:
$$
T_{obs} = \frac{456.83624 - 0}{27.70037} = 16.492 
$$
The $P$-value is $P(T_{14 - 2} > 16.492) = 0.000000000654037$. This is no surprise, given our previous finding of $396.4823 \leq A \leq 517.1902$.
```{r}
options(scipen=999)
1 - pt(16.492, 14-2)
```
</br>
</br>

**Demonstration 2B:** Does this data support the notation that a $Y$-intercept term in the model that attempts to predict the average SAT Math score of all students in a certain state as a linear function of the state's average teacher salary for all states where the proportion of students who write the SAT is **high** is less than 500?  Test using a $P({\rm Type}\:\:I\:\:{\rm Error}) = 0.05$.
</br>
</br>
**Answer:** We wish to test the following

$$
{\rm H}_{0}:A = (\geq)\:\:500  \hspace{0.5in} {\rm H}_{A}: A <  500 
$$


Assuming the null hypothesis is true, we presume that the value of $A = 600$. We recall that the value of $a = 456.83624$ and $S_{a} = 27.70037$. The test statistic $T_{obs}$ is computed

$$
T_{obs} = \frac{456.83624 - 500}{27.70037} = -1.5582
$$



The $P$-value is $P(T_{14 - 2} < -1.5582) = 0.07258$ which is computed with R

```{r}
pt(-1.5582, 14-2)
```
</br>
</br>
</br>


### 2B. Inferences About $B$

#### Confidence Interval Estimation of $B$

Similar to the confidence interval estimation of $A$, the confidence interval estimate for $B$ is of the same form:
$$
b \pm \left(t_{0.975,n-2}*S_{b}\right)
$$


With respect to our problem, the value of $S_{b}$ can also be extracted from the `coefficients(summary(modelname))`:
```{r}
coefficients(summary(predictmathSAT))
```

Notice the value beneath the "Std.Error" column associated with "salary", that is the value of $S_{b} = 0.000452$. As a result, the 95% confidence interval for $B$ is 
$$
b \pm \left(t_{0.975,n-2}*S_{b}\right) \longrightarrow 0.00076 \pm (2.178813*0.00045) \longrightarrow 0.00076 \pm 0.00098 \longrightarrow -0.0002 \leq B \leq 0.0020
$$
This result can be obtained from
```{r}
confint(predictmathSAT) #refer to the second row for a 95% CI for B
```
Subsequently, a 90% confidence interval for $B$ (the last of the two) and $A$ (the first of the two) is
```{r}
confint(predictmathSAT, level=0.90)
```

#### Hypothesis Testing about the value of $B$

There are three different statistical hypotheses that can be carried out on the value of the $B$, or slope, term:
$$
\begin{array}{lccc}
{\rm Right-Tailed}                   & {\rm Left-Tailed}       & {\rm Two-Tailed}  \\
{\rm H}_{0}: B \leq B^{*}  & {\rm H}_{0}: B \geq B^{*}  & {\rm H}_{0}: B = B^{*} \\
{\rm H}_{A}: B > B^{*}     & {\rm H}_{A}: B < B^{*}     & {\rm H}_{A}: B \ne B^{*}
\end{array}
$$
The test statistic is computed with $T_{obs}$
$$
T_{obs} = \frac{b - B^{*}}{S_{b}} \sim T_{n -2}
$$
</br>
**Demonstration 1:** Consider the statistical hypothesis 
$$
{\rm H}_{0}: B = 0  \:\: \text{(Y cannot be expressed as a linear function of X)} \\
{\rm H}_{A}: B \ne 0 \:\: \text{(Y can be expressed as a linear function of X)} 
$$
The results of the $F$-test were
```{r}
summary(aov(predictmathSAT))
```

How does the $t$-test compare?
</br>
</br>

$$
T_{obs} = \frac{b - B^{*}}{S_{b}} = \frac{0.00076 - 0}{0.00045} = 1.6849
$$
The $P$-value is $P(T_{14-2} > 1.6849)*2 = 0.1178149 \approx 0.118$. 
```{r}
(1 - pt(1.6849, 14-2))*2
```
</br>

--------------------

**A comment about the table of coefficients**. The results of ${\rm H}_{0}: A = 0$ and ${\rm H}_{0}: B = 0$ compared to ${\rm  H}_{A}: A \ne 0$ and ${\rm H}_{A}: B \ne 0$ can be obtained from
```{r}
coefficients(summary(predictmathSAT))
```
</br>

---------------------

**Time to Play 1:** Recall from Thursday's class, your attempt(s) to model an American League baseball team's home attendance (**seasonattend**) as a linear function of the team's winning percentage (**winningpct**). You may have to re-run Thursday's code to read in the data file and create these two variables. 

(a) Compute the 95% confidence inteval for the $A$-term of the model. 
</br>
</br>
**Answer:**
```{r}
MLBattend <- read.csv("http://people.ucalgary.ca/~jbstall/DataFiles/baseball.csv")
MLBattend <- MLBattend %>%
  mutate(winningpct = c(wins / (wins + losses)))
MLBattend = MLBattend %>% 
  mutate(seasonattend = (attendance/1000000))
predictattendanceAL = lm(seasonattend~winningpct, data = filter(MLBattend, league == "AL"))
confint(predictattendanceAL)
```

$$
-1.014 \leq A \leq -0.087
$$
In this instance, the value of $A$ represents the average/mean home season attendance for an American League team that has a winning percentage of 0.00.


<div style="margin-bottom:200px;">
</div>

(b) What does the interval in (a) represent, that is, what does your finding mean with respect to these data?
</br>
</br>
**Answer:**
In this instance, the value of $A$ represents the average/mean home season attendance for an AL team that has a winning percentage of 0.00.


<div style="margin-bottom:100px;">
</div>

(c) Can you infer that an American League baseball team's home attendance can be expressed a positive linear function of the team's winning percentage? State your statistical hypotheses, find/compute the value of the test statistic, and find/compute the $P$-value. 
</br>
</br>
**Answer:**
If we can express an American League baseball team's home attendance as a *positive* linear function of their winning percentage, then
$B > 0$. So, we test
$$
{\rm H}_{0}: B = \:\:(\text{or}\:\:\leq) \:\: 0 \hspace{0.5in} {\rm H}_{A}: B > 0
$$
The value of the test statistic $T_{obs}$ can be extracted from R
```{r}
coefficients(summary(predictattendenceAL))
```
from which
$$
T_{obs} = \frac{4.5352 - 0}{0.4672} = 9.705
$$
The $P$-value is $P(T_{432-2} > 9.705) = \frac{0.0000000000000000000288423}{2}$.
```{r}
1-pt(9.705, 432 - 2)
```
The null hypothesis is rejected. One can infer from these data that a American League baseball team's home attendance *can be* expressed as a positive linear function of their season winning percentage.

<div style="margin-bottom:200px;">
</div>

(d) Based on your result in part(c), compute the 95% confidence interval for $B$. Also, interpret its meaning in the context of these data.
</br>
</br>
**Answer:**
Refer to the output in part (a), the 95% confidence interval for $B$ is
$$
3.616737 \leq B \leq   5.45376658
$$
If a team's winning percentage increases by 1% (in this case, 0.01), then a team's season attendance will **increase by an average** of anywhere between
$$
3.616737*(0.01) \leq B \leq   5.45376658*(0.01)  \longrightarrow 0.036 \:\:\text{to}\:\; 0.0545
$$
That is, as the winning percentage increases by 1% (0.01), this will result in an *average increase* in season attendance from about 36,000 to 54,500 persons. 


<div style="margin-bottom:200px;">
</div>


----------------------

## What if the Statistical Result is Significant - Computing Confidence Intervals for the $y$-Variable for a Given value of the $x$-Variable

For a certain value of the $x$-variable/predictor variable, one can take into account the "random behaviour" of the estimate of the model to "add certainty" to the prediction of the $y$-variable/response variable.

Earlier, we saw how we could predict the average value of the $y$-variable for a certain value of the $x$-variable. 

For example, a certain state's (with a "high" percentage of students writing the SAT exam in 2013) average teacher salary is $55,000. The predicted/average SAT math score for *all such state's* is
```{r}
predict(predictmathSAT, newdata=data.frame(salary=55000))
```
and
$$
\widehat{y}_{i} = 456.83624 + (0.000763*55000) = 498.777 \approx 498.78
$$

We can broaden this prediction to

1. predict the mean value of the $y$-variable, $\mu_{Y|x}$, the mean of a sub-population of $y$-values for a certain value of the $x$-variable

2. predict a single, or the "next" $y$-value, $Y|x$, for a sub-population of $y$-values for a given value of the $x$-variable

This can be done with an added "level of certainty" through the $100(1-\alpha)$% confidence interval estimate of either 

1. $\mu_{Y|x}$

2. $Y|x$


#### 1.  Confidence Interval Estimation of $\mu_{y|x}$


**Result:** A $100(1 - \alpha)$% confidence interval estimate for the mean value of $Y$ for a given value of $X = x_{p}$ is 

$$
(a + b*x_{p})  \pm  t_{\frac{\alpha}{2}, n - 2} (S_{e}) \sqrt{\frac{1}{n} + \frac{(x_{p} - \overline{X})^{2}}{\sum X^{2} - n(\overline{X})^{2}}}
$$
To obtain a confidence interval for the mean value of $y$ for a given value of $x$, we can add the **interval="conf"** option to the `predict` command

```{r}
predict(predictmathSAT, newdata=data.frame(salary=55000), interval="conf", conf.level=0.95)
```
Including the **"interval=conf"** option instructs R to compute the 95% confidence inteval for the mean "average SAT math score" for all states that have an average teacher salary of $x_{p} = 55000$ ,and
$$
488.8271 \leq \mu_{Y|x = 55000} \leq 508.7269
$$
</br>
</br>

#### 2. Capturing $Y|_{X = x_{p}}$ Via Confidence Interval Estimation

Compared to confidence interval estimation of $\mu_{Y|X=x_{p}}$, we can also find a confidence interval for a *single* value of $Y$ for a given value of the response/explanatory variable $X = x_{p}$. The formula to do so is

$$
(a + b*x_{p})  \pm  t_{\frac{\alpha}{2}, n - 2} (S_{e}) \sqrt{1 +\frac{1}{n} + \frac{(x_{p} - \overline{X})^{2}}{\sum X^{2} - n(\overline{X})^{2}}}
$$

to obtain this interval, a prediction interval for $Y$ for a given value of $X = x_{p}$, change the interval option to **interval="predict"**:
```{r}
predict(predictmathSAT, newdata=data.frame(salary=55000), interval="predict", conf.level=0.95)
```
$$
466.3227 \leq Y|x = 55000 \leq 531.2312
$$
</br>

**Time to Play 2A:** The Boston Red Sox (an American League team) had a winning percentage this year (2019) of 0.519. With 95% confidence, predict the total home attendance for the Boston Red Sox for 2019. Interpret your result. 
</br>
</br>
**Answer:**
We are estimating/predicting the home season attendance for *one* team with a winning percentage of $x = 0.519$. 
```{r}
predict(predictattendenceAL, newdata=data.frame(winningpct=0.519), interval="predict")
```
from which
$$
0.4766774 \leq Y|x= 0.519 \leq 3.129839 \:\: \text{(millions of people)}
$$

<div style="margin-bottom:200px;">
</div>


**Time to Play 2B:** The Oakland A's had a winning percentage of 0.600. With 98% confidence, what is the expected home attendance of an American League team with a winning percentage of 0.600? Compute the interval and interpret your result. 
</br>
</br>
**Answer:**
We are estimating/predicting mean/average home season attendance for *all* AL teams with a winning percentage of $x = 0.600$. 
```{r}
predict(predictattendenceAL, newdata=data.frame(winningpct=0.600), interval="conf", level=0.98)
```
from which
$$
2.037634 \leq \mu_{Y|x= 0.600} \leq 2.303589 \:\: \text{(millions of people)}
$$

<div style="margin-bottom:200px;">
</div>