---
title: "Inference of contingency table"
output: html_notebook
---

##Two-way contingency table test
```{r}
# explore the origin of cars and their prices
library(MASS)
data(Cars93)
M_price<-median(Cars93$Price)
L1=sum(Cars93$Price<=M_price)
mydata<-Cars93[order(Cars93$Price),]
Yes<-c(rep(0,L1),rep(1,(length(Cars93$Price)-L1)))
carDATA<-cbind(Yes,mydata)
```

```{r}
tab<-table(carDATA$Origin, carDATA$Yes)
tab
```

**Test of difference**
```{r}
library(fmsb)
tab_M<-as.matrix(tab)
# function "riskdifference" tests the difference of proportions
risk_difference<-riskdifference(27,20,48,45, conf.level = 0.95)
risk_difference
```

```{r}
# do it manually
pi1_hat=27/48
pi2_hat=20/45
var_diff=pi1_hat*(1-pi1_hat)/48+pi2_hat*(1-pi2_hat)/45
# the estimated difference
esti_diff=pi1_hat-pi2_hat
# the 95 percent CI
print(c(esti_diff-1.96*sqrt(var_diff), esti_diff+1.96*sqrt(var_diff)))

```

**Test the ratio (relative risk)**
```{r}
# function "riskratio" tests the relative risk
risk_ratio<-riskratio(27,20,48,45, conf.level = 0.95, p.calc.by.independence = TRUE)
risk_ratio
```

```{r}
RR<-pi1_hat/pi2_hat
var_logRR<-(1-pi1_hat)/27+(1-pi2_hat)/20
# the 95% CI for log RR
print(c(log(RR)-1.96*sqrt(var_logRR), log(RR)+1.96*sqrt(var_logRR)))
```
```{r}
# the 95% CI for RR
print(c(exp(log(RR)-1.96*sqrt(var_logRR)), exp(log(RR)+1.96*sqrt(var_logRR))))
```

**Test the odds ratio**
```{r}
# function "oddsratio" tests the odds ratio
test_odds<-oddsratio(tab_M, conf.level = 0.95, p.calc.by.independence = TRUE)
test_odds
```

```{r}
esti_odds_ratio=pi1_hat*(1-pi2_hat)/pi2_hat/(1-pi1_hat)
var_odds_ratio=1/27+1/21+1/20+1/25
# the 95% CI for log odds_ratio is
print(c(log(esti_odds_ratio)-1.96*sqrt(var_odds_ratio), log(esti_odds_ratio)+1.96*sqrt(var_odds_ratio)))
```

```{r}
# the 95% CI for odds ratio is
print(c(exp(log(esti_odds_ratio)-1.96*sqrt(var_odds_ratio)),exp(log(esti_odds_ratio)+1.96*sqrt(var_odds_ratio))))
```

_______________________________________
**In-class practice question 1**
```{r}
# explore whether the price is related to the manual transmission cars
carDATA<-Cars93[order(Cars93$Price),]
l1=sum(as.numeric(carDATA$Price)<=median(carDATA$Price))
l2=dim(carDATA)[1]-l1
Label=c(rep('Cheap',l1),rep('Expensive',l2))
newDATA<-cbind(Label,carDATA)
tab2<-table(newDATA$Label, newDATA$Man.trans.avail)
tab2
```
```{r}
# use risk difference test
risk_difference<-riskdifference(9,23,47,46, conf.level = 0.95)
risk_difference
```

```{r}
# use relative risk test (risk ratio)
risk_ratio<-riskratio(9,23,47,46, conf.level = 0.95, p.calc.by.independence = TRUE)
risk_ratio
```

```{r}
# use odds ratio test
test_odds<-oddsratio(tab2, conf.level = 0.95, p.calc.by.independence = TRUE)
test_odds
```

______________________________________
##An example from the lecture notes
```{r}
MF_politics<-matrix(c(762,327,468,484,239,477), nrow = 2, byrow = TRUE)
MF_tab<-as.table(MF_politics)
chisq.test(MF_tab)
```

_______________________________________
##Chi-square multinomial test
```{r}
C_summary<-summary(Cars93$Horsepower)
carDATA<-Cars93[order(Cars93$Horsepower),]
l1=sum(as.numeric(carDATA$Horsepower<=C_summary[2]))
l2=sum(as.numeric(carDATA$Horsepower<=C_summary[3]))-l1
l3=sum(as.numeric(carDATA$Horsepower<=C_summary[4]))-l1-l2
l4=dim(carDATA)[1]-l1-l2-l3
Horse_level<-c(rep('Level 0',l1),rep('Level 1',l2),rep('Level 2',l3), rep('Level 3', l4))
newDATA<-cbind(Horse_level,carDATA)
tab3<-table(newDATA$Cylinders, newDATA$Horse_level)
tab3
```

```{r}
chisq.test(tab3)
```

```{r}
ctab3<-rbind(tab3[1,]+tab3[2,],tab3[3,]+tab3[4,], tab3[5,]+tab3[6,])
ctab3
```

```{r}
cctab3<-cbind(ctab3[,1],ctab3[,2],ctab3[,3]+ctab3[,4])
cctab3
```

```{r}
chisq.test(cctab3)
```

_______________________________________
**In-class practice question 2**
```{r}
# explore the relationship between smoking habit and physical exercise situation
# Please run chi-square test to see if they are independent
# If you got the warning message, please combine the last two columns and redo the chi-square test
tbl = table(survey$Smoke, survey$Exer) 
tbl
```

```{r}
# Your answer
chisq.test(tbl)
```
```{r}
tbl2<-data.frame(tab[,1],tbl[,2]+tbl[,3])
colnames(tbl2)=c('Freq','None or some')
rownames(tbl2)=c('Heavy','Never','Occas','Regul')
chisq.test(tbl2)
```
________________________________________
**In-class practice question 3**
```{r}
# Establish the table by yourself to test whether "DriveTrain" is related to "Cylinders" (using Cars93 dataset)
tbl3<-table(Cars93$DriveTrain, Cars93$Cylinders)
chisq.test(tbl3)
```

_________________________________________
## Pearson residuals 
```{r}
MF_politics<-matrix(c(762,327,468,484,239,477), nrow = 2, byrow = TRUE)
MF_tab<-as.table(MF_politics)
library(questionr)
chisq.residuals(MF_tab, std=TRUE)
```

```{r}
# survival data of Titanic
data(Titanic)
Ti_tab<-apply(Titanic, c(1,4), sum)
Ti_tab
```

__________________________________________
**In-practice question 4**
```{r}
# Test whether travel class and survivial rate are dependent
chisq.test(Ti_tab)
```

```{r}
# If they are dependent, which cell deviates most from the independence assumption
chisq.residuals(Ti_tab, std=TRUE)
```

___________________________________________
## Ordinal data test
```{r}
AL_Infant<-matrix(c(48,17066,38,14464,5,788,1,126,1,37),nrow = 5,byrow = TRUE)
AL_tab<-as.table(AL_Infant)
AL_tab
```

```{r}
# A self-defined function to calculate the Mantel-Haenszel statistic, as well as the p-value
pears.cor=function(table, rscore, cscore)
{ 
	dim=dim(table) 
	rbar=sum(margin.table(table,1)*rscore)/sum(table) 
	rdif=rscore-rbar 
	cbar=sum(margin.table(table,2)*cscore)/sum(table) 
	cdif=cscore-cbar 
	ssr=sum(margin.table(table,1)*(rdif^2)) 
	ssc=sum(margin.table(table,2)*(cdif^2)) 
	ssrc=sum(t(table*rdif)*cdif) 
	pcor=ssrc/(sqrt(ssr*ssc)) 
	pcor 
	M2=(sum(table)-1)*pcor^2
	M2
	result=c(pcor, M2, (1-pchisq(M2,1)))
	result=as.table(result)
	names(result)=c('Pearson correlation','MH statistic', 'P-Value')
	result
} 
```

```{r}
# Pearson chi-square test
chisq.test(AL_tab)
```

```{r}
pears.cor(AL_tab,c(0,0.5,1.5,4,7),c(0,1))
```

```{r}
# try difference scores
pears.cor(AL_tab,c(0,0.5,1.5,4,7),c(1,3))
```

______________________________________________
**In-class practice 5: Titanic data again**
```{r}
# Treat the variables as ordinal and apply Mantel-Haenszel test
data(Titanic)
Ti_tab<-apply(Titanic, c(1,4), sum)
Ti_tab
```

```{r}
# When you assgin scores to row and column variables, try different scores to check the result.
pears.cor(Ti_tab, c(1,2,3,4), c(0,1))
```
```{r}
pears.cor(Ti_tab, c(1,3,5,7), c(0,1))
```

______________________________________________
## nominal-ordinal data
```{r}
library(MASS)
data(Cars93)
car_tab<-table(Cars93$Type, Cars93$Cylinders)
car_tab<-car_tab[,c(1,2,3,4,5)]
car_tab
```

```{r}
# suppose you are given such a table, how to convert it back to the data.frame
# step 1: give names to dimensions
names(dimnames(car_tab))=c('Type','Cylinders')
car_tab
```

```{r}
# step 2: convert the table to data frame
tab_car<-as.data.frame(car_tab)
tab_car
```

```{r}
# step 3: recover the original data frame
Type=vector()
Cylinders=vector()

# use "as.character" function to convert factor to character, such as Type variable: "Compact", "Large", etc.
# use "as.numeric" and "levels" functions to convert factor to numbers, such as Cylinders variable: "3", "4", etc.
for (i in 1:length(tab_car$Freq)){
  Type=c(rep(as.character(tab_car$Type[i]), tab_car$Freq[i]), Type)
  Cylinders=c(rep(as.numeric(levels(tab_car$Cylinders[i]))[tab_car$Cylinders[i]], tab_car$Freq[i]), Cylinders)
}

tab_car2<-data.frame(Type, Cylinders)
tab_car2
```

```{r}
# Step 4: use function "aov" to do ANOVA 
car_aov<-aov(Cylinders~Type, data = tab_car2)
summary(car_aov)
```

______________________________________
**In-class practice 6**
```{r}
# use above method to test whether "DriveTrain" is related to "Cylinders"
# Use the table given below
Dr_Cy_tab<-table(Cars93$DriveTrain, Cars93$Cylinders)
Dr_Cy_tab<-Dr_Cy_tab[,c(1,2,3,4,5)]
Dr_Cy_tab
```

```{r}
# Your answer
names(dimnames(Dr_Cy_tab))=c('DriveTrain','Cylinders')
Dr_Cy_tab<-as.data.frame(Dr_Cy_tab)

Dr=vector()
Cy=vector()

# use "as.character" function to convert factor to character, such as Type variable: "Compact", "Large", etc.
# use "as.numeric" and "levels" functions to convert factor to numbers, such as Cylinders variable: "3", "4", etc.
for (i in 1:length(Dr_Cy_tab$Freq)){
  Dr=c(rep(as.character(Dr_Cy_tab$DriveTrain[i]), Dr_Cy_tab$Freq[i]), Dr)
  Cy=c(rep(as.numeric(levels(Dr_Cy_tab$Cylinders[i]))[Dr_Cy_tab$Cylinders[i]], Dr_Cy_tab$Freq[i]), Cy)
}

tab_car3<-data.frame(Dr, Cy)
tab_car3
```
```{r}
car3_aov<-aov(Cy~Dr, data = tab_car3)
summary(car3_aov)
```


_______________________________________
## Fisher's exact test
```{r}
TeaTasting <-
matrix(c(3, 1, 1, 3),
       nrow = 2,
       dimnames = list(Truth = c("Milk", "Tea"),Guess = c("Milk", "Tea")))
TeaTasting
```

```{r}
# DO remember
fisher.test(TeaTasting, alternative = "two.sided")
```

**In-class practice 7**
```{r}
# practice 6 revisited
Dr_Cy_tab<-table(Cars93$DriveTrain, Cars93$Cylinders)
Dr_Cy_tab<-Dr_Cy_tab[,c(1,2,3,4,5)]
Dr_Cy_tab
```

```{r}
# Use Pearson Chi-square test to check whether "DriveTrain" is related to "Cylinders"
chisq.test(table(Cars93$DriveTrain, Cars93$Cylinders))
```

```{r}
# Did Chi-square test give you accurate test result?
# If not, try fisher's exact test
fisher.test(table(Cars93$DriveTrain, Cars93$Cylinders), alternative = "two.sided")
```