---
title: "Introduction to contingency table"
output: html_notebook
---

```{r}
library(MASS)
data(Cars93)
Cars93[1:15,]
```

```{r}
# Get a two-way contingency table (joint table)
tab1<-table(Cars93$Origin, Cars93$Type)
tab1
```

```{r}
# get the proportions in the joint table
prop.table(tab1)
```

```{r}
# get the row sums
rowSums(tab1)
margin.table(tab1,1)
```

```{r}
# get the column sums
colSums(tab1)
margin.table(tab1,2)
```

```{r}
# get the proportions in conditional table
prop.table(tab1, margin=1)
```

**Question: could you give a two-way contingency table of "Type" and "Cylinders"? Calculate the row and column sums as well the proportion conditional table**
```{r}
# Your answer
tabb<-table(Cars93$Type, Cars93$Cylinders)
margin.table(tabb,1)
margin.table(tabb,2)
prop.table(tabb, margin=1)
```


```{r}
# A three-way contingency (joint) table
tab2<-table(Cars93$Origin, Cars93$Type, Cars93$Cylinders)
ftable(tab2)
```

```{r}
ftable(prop.table(tab2))
```

```{r}
ftable(prop.table(tab2,margin = 2))
```

```{r}
ftable(prop.table(tab2,margin = c(1,2)))
```

```{r}
margin.table(tab2, margin = c(1,2))
```

**Question: get the marginal table of "Origin" and "Cylinders"?**
```{r}
# Your answer
margin.table(tab2, margin = c(1,3))
```


**Now we have another table, which is constructed using Cars93**
```{r}
library(MASS)
data(Cars93)
M_price<-median(Cars93$Price)
L1=sum(Cars93$Price<=M_price)
mydata<-Cars93[order(Cars93$Price),]
Yes<-c(rep(0,L1),rep(1,(length(Cars93$Price)-L1)))
carDATA<-cbind(Yes,mydata)
```

```{r}
# 0 means cheap car; 1 means luxury car
tab<-table(carDATA$Origin, carDATA$Yes)
tab
```

**Question: to check whether the origin and price are relavent, calculate the difference of proportions, relative risk and odds ratio between US cars and Non-US cars**
```{r}
tab_prop<-prop.table(tab,margin=1)
tab_prop
```
```{r}
# difference of proportions
tab_prop[1,1]-tab_prop[2,1]
```

```{r}
# relative risk
tab_prop[1,1]/tab_prop[2,1]
```

```{r}
# odds ratio
tab_prop[1,1]*tab_prop[2,2]/(tab_prop[2,1]*tab_prop[1,2])
```