---
title: "DATA 602 Project: Exploring the City of Calgary's Public Trees Dataset"
author: Harneet Cheema, Michael Ellsworth
output:
  ioslides_presentation: default
---

## Introduction
- Calgary has ~7 million trees
- Calgary should not have ~7 million trees
- Maintaining Public Trees is a costly endeavour
- How could Calgary's tree population improve?
- What insights can we gain from public tree data?

## Dataset
- Calgary has data on ~500,000 trees
- 20 different features
- Analysis on tree condition

    - Common Name / Genus / Size
  
- Analysis on planting trends

    - Proportions in Developing / Developed Communities
  
## Packages
- dplyr
- ggplot2
- readr
- stringr
- mosaic

```{r Load Packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(mosaic)
```

```{r Load Data, include=FALSE}
#trees <- read_csv("https://data.calgary.ca/api/views/tfs4-3wwa/rows.csv?accessType=DOWNLOAD")
#comm_info <- read_csv("https://data.calgary.ca/api/views/j9ps-fyst/rows.csv?accessType=DOWNLOAD")
trees <- read_csv("Public_Trees.csv")
comm_info <- read_csv("City_of_Calgary_Community_Areas.csv")
```

```{r Clean Data, include=FALSE}
# Change column names in community info to upper case
names(comm_info) <- toupper(names(comm_info))

# Select columns to keep in each dataset
trees_clean <- trees %>%
  select(DECIDUOUS_EVERGREEN,
         COMMON_NAME,
         GENUS,
         SPECIES,
         DBH_CM,
         MATURE_SIZE,
         RATING,
         TREE_CONDITION_RATING_PERC,
         HERITAGE_TREES,
         COMM_CODE)

comm_info_clean <- comm_info %>%
  select(CLASS,
         COMM_CODE,
         NAME,
         SECTOR,
         SRG)

# Merge community information and trees
public_trees <- inner_join(trees_clean, comm_info_clean, by = "COMM_CODE")

# Convert all character strings to upper case
public_trees$DECIDUOUS_EVERGREEN <- str_to_upper(public_trees$DECIDUOUS_EVERGREEN)
public_trees$COMMON_NAME <- str_to_upper(public_trees$COMMON_NAME)
public_trees$GENUS <- str_to_upper(public_trees$GENUS)
public_trees$SPECIES <- str_to_upper(public_trees$SPECIES)
public_trees$MATURE_SIZE <- str_to_upper(public_trees$MATURE_SIZE)
public_trees$CLASS <- str_to_upper(public_trees$CLASS)
```

## Explore the data
- Many different features to look at
- Determine the most popular trees by name / genus

    - Isolate the analysis on these types of trees
    
- Tree size

## Explore - Common Tree Names
- Top 10 Trees by Common Name in Calgary
- Select "Green Ash"

```{r List Top 10 Names, include=FALSE}
# Tally number of trees per tree name and plot
tree_names <- public_trees %>%
  filter(!is.na(COMMON_NAME)) %>% # Filters NAs from species column
  group_by(COMMON_NAME) %>%
  tally() %>%
  arrange(desc(n))
names_top10 <- pull(tree_names %>% top_n(10), COMMON_NAME)
```

```{r Top 10 Trees, echo=FALSE, warning=FALSE}
tree_names_plot <- tree_names %>%
  filter(COMMON_NAME %in% names_top10) %>%
  ggplot(aes(x = reorder(COMMON_NAME, n), y = n / 1000)) +
  geom_col() +
  xlab("Tree Name") +
  ylab("Total Count (Thousands) - Calgary") +
  ggtitle("Tree Count By Common Name - Top 10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
tree_names_plot
```

## Explore - Genus
- Top 10 Trees by Genus in Calgary
- Select "Picea"

```{r List Top 10 Genus, include=FALSE}
# Tally number of trees per Genus type and plot
genus <- public_trees %>%
  filter(!is.na(GENUS)) %>% # Filters NAs from species column
  group_by(GENUS) %>%
  tally() %>%
  arrange(desc(n))
genus_top10 <- pull(genus %>% top_n(10), GENUS)
```

```{r Top 10 Genus, echo=FALSE, warning=FALSE}
genus_plot <- genus %>%
  filter(GENUS %in% genus_top10) %>%
  ggplot(aes(x = reorder(GENUS, n), y = n / 1000)) +
  geom_col() +
  xlab("Tree Genus") +
  ylab("Total Count (Thousands) - Calgary") +
  ggtitle("Tree Count By Genus - Top 10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
genus_plot
```

## Explore - Size
- Large trees are most common, small are least common
- For comparison, Large and Small trees are selected
```{r Explore Size, include=FALSE}
# Tally number of trees per Genus type and plot
size <- public_trees %>%
  filter(!is.na(MATURE_SIZE)) %>% # Filters NAs from species column
  group_by(MATURE_SIZE) %>%
  tally() %>%
  arrange(desc(n))
```

```{r Size Plot, echo=FALSE, warning=FALSE}
size_plot <- size %>%
  ggplot(aes(x = reorder(MATURE_SIZE, n), y = n / 1000)) +
  geom_col() +
  xlab("Tree Size") +
  ylab("Total Count (Thousands) - Calgary") +
  ggtitle("Tree Count By Size") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
size_plot
```

## Difference of Tree Condition Means - Green Ash and Other
- Test how average Green Ash condition compares against all other trees
- Is there a reason why this is the most popular tree in Calgary?
- Should it be the most popular tree in Calgary?

## Difference of Tree Condition Means - Green Ash and Other
$$
\begin{array}{lcc}
               &   {\rm Stastical \ Hypothesis}\\
{\rm H}_{0}: & \mu_{\text{Tree Condition}(Green Ash)} \leq \mu_{\text{Tree Condition}(Other)}       \\
{\rm H}_{A}: & \mu_{\text{Tree Condition}(Green Ash)}  > \mu_{\text{Tree Condition}(Other)} 
\end{array}
$$

- Statistical Hypothesis is based on assumption that the most popular tree would have a better overall condition than other trees

## Difference of Tree Condition Means - Green Ash and Other
- Convential confidence interval estimation of the difference between two population means
- Conditions can be relaxed since n = 450,632
```{r echo=FALSE}
# T-Test Dataframe (two parameter - Green ash versus other)
# Create a Data Frame for all Tree Condition values EXCEPT Green Ash (name them OTHER)
total_conditions_name <- public_trees %>%
  filter(!is.na(COMMON_NAME), !is.na(TREE_CONDITION_RATING_PERC)) %>%
  filter(TREE_CONDITION_RATING_PERC <= 100) %>%
  filter(COMMON_NAME != "GREEN ASH") # Keeps all trees except Green Ash
total_conditions_name$COMMON_NAME <- "OTHER" # Names all trees except Green Ash "OTHER"

# Create a Data Frame for all Tree Condition values for Green Ash
GA_conditions <- public_trees %>%
  filter(!is.na(COMMON_NAME), !is.na(TREE_CONDITION_RATING_PERC)) %>%
  filter(TREE_CONDITION_RATING_PERC <= 100) %>%
  filter(COMMON_NAME == "GREEN ASH")

# Bind the two data frames together and select only Condition and Name columns
GA_and_total_conditions <- rbind(GA_conditions, total_conditions_name)
GA_and_total_conditions <- GA_and_total_conditions %>%
  select(TREE_CONDITION_RATING_PERC, COMMON_NAME)

# T-Test - two parameter (Green Ash mean = Other mean)
t.test(~ TREE_CONDITION_RATING_PERC | COMMON_NAME,
       conf.level = 0.95,
       alternative = "greater",
       data = GA_and_total_conditions)
```

## Difference of Tree Condition Means - Green Ash and Other
- The P-Value in the t-test is 1
- Cannot reject our null hypothesis
- Average tree condition of other trees is greater than average condition of Green Ash
- Should Green Ash be the most popular tree if the condition doesn't out perform the others?

```{r include=FALSE}
# Checks for the normality condition although this condition can be relaxed since n > 25
ggplot(data=GA_conditions, aes(sample = TREE_CONDITION_RATING_PERC)) + stat_qq(size=2, col='blue') + stat_qq_line(col='red')
ggplot(data=total_conditions_name, aes(sample = TREE_CONDITION_RATING_PERC)) + stat_qq(size=2, col='blue') + stat_qq_line(col='red')
```

## Difference of Tree Condition Means - Picea and Other
- Test how average Picea condition compares against all other trees
- Is there a reason why this is the most popular tree Genus in Calgary?
- Should it be the most popular tree Genus in Calgary?

## Difference of Tree Condition Means - Picea and Other
$$
\begin{array}{lcc}
               &   {\rm Stastical \ Hypothesis}\\
{\rm H}_{0}: & \mu_{\text{Tree Condition}(Picea)} \leq \mu_{\text{Tree Condition}(Other)}       \\
{\rm H}_{A}: & \mu_{\text{Tree Condition}(Picea)}  > \mu_{\text{Tree Condition}(Other)} 
\end{array}
$$

- Statistical Hypothesis is based on assumption that the most popular tree would have a better overall condition than other trees

## Difference of Tree Condition Means - Picea and Other
- Convential confidence interval estimation of the difference between two population means
- Conditions can be relaxed since n = 450,632
```{r echo=FALSE}
# T-Test Dataframe (two parameter - Picea versus other)
# Create a Data Frame for all Tree Condition values EXCEPT Picea (name them OTHER)
total_conditions_genus <- public_trees %>%
  filter(!is.na(GENUS), !is.na(TREE_CONDITION_RATING_PERC)) %>%
  filter(TREE_CONDITION_RATING_PERC <= 100) %>%
  filter(GENUS != "PICEA") # Keeps all trees except Picea
total_conditions_genus$GENUS <- "OTHER" # Names all trees except Picea "OTHER"

# Create a Data Frame for all Tree Condition values for Picea
picea_conditions <- public_trees %>%
  filter(!is.na(GENUS), !is.na(TREE_CONDITION_RATING_PERC)) %>%
  filter(TREE_CONDITION_RATING_PERC <= 100) %>%
  filter(GENUS == "PICEA")

# Bind the two data frames together and select only Condition and Name columns
picea_and_total_conditions <- rbind(picea_conditions, total_conditions_genus)
picea_and_total_conditions <- picea_and_total_conditions %>%
  select(TREE_CONDITION_RATING_PERC, GENUS)

# T-Test - two parameter (Picea mean = Other mean)
t.test(~ TREE_CONDITION_RATING_PERC | GENUS,
       conf.level = 0.95,
       alternative = "less",
       data = picea_and_total_conditions)
```

## Difference of Tree Condition Means - Picea and Other
- The P-Value in the t-test is less than 0.05
- Reject our null hypothesis
- Average tree condition of Picea is greater than average condition of other trees
- Picea appear to be a good pick for the most popular tree Genus in Calgary!

## Difference of Tree Condition Means - Large and Small trees
- Test how average tree condition compares between Large and Small trees
- Should Calgary be planting more Small trees?

## Difference of Tree Condition Means - Large and Small trees
$$
\begin{array}{lcc}
               &   {\rm Stastical \ Hypothesis}\\
{\rm H}_{0}: & \mu_{\text{Tree Condition}(Large)} \geq \mu_{\text{Tree Condition}(Small)}       \\
{\rm H}_{A}: & \mu_{\text{Tree Condition}(Large)}  < \mu_{\text{Tree Condition}(Small)} 
\end{array}
$$

- Statistical Hypothesis is based on our assumption that large trees would have lower tree condition because Calgary isn't a great climate for trees

## Difference of Tree Condition Means - Large and Small trees
```{r echo=FALSE}
# T-Test Dataframe (two parameter - Large versus Small)
# Create a Data Frame for all Tree Condition values for Large trees
large_conditions <- public_trees %>%
  filter(!is.na(MATURE_SIZE), !is.na(TREE_CONDITION_RATING_PERC)) %>%
  filter(TREE_CONDITION_RATING_PERC <= 100) %>%
  filter(MATURE_SIZE == "LARGE") # Keeps all Large

# Create a Data Frame for all Tree Condition values for Small trees
small_conditions <- public_trees %>%
  filter(!is.na(GENUS), !is.na(TREE_CONDITION_RATING_PERC)) %>%
  filter(TREE_CONDITION_RATING_PERC <= 100) %>%
  filter(MATURE_SIZE == "SMALL")

# Bind the two data frames together and select only Condition and Name columns
large_and_small_conditions <- rbind(large_conditions, small_conditions)
large_and_small_conditions <- large_and_small_conditions %>%
  select(TREE_CONDITION_RATING_PERC, MATURE_SIZE)

# T-Test - two parameter (Picea mean = Other mean)
t.test(~ TREE_CONDITION_RATING_PERC | MATURE_SIZE,
       conf.level = 0.95,
       alternative = "less",
       data = large_and_small_conditions)
```

## Difference of Tree Condition Means - Large and Small trees
- The P-Value in the t-test is 0.0297
- Reject our null hypothesis
- Average tree condition of small trees is greater than average condition of large trees
- Should small trees be planted more often than large trees?
- The mean values are very close - is it negligible?

## Tree Condition vs Size (Diameter of Trunk)
```{r echo=FALSE}
# Tree Condition vs Diameter of Trunk
public_trees %>%
  filter(!is.na(TREE_CONDITION_RATING_PERC), !is.na(DBH_CM)) %>%
  filter(TREE_CONDITION_RATING_PERC < 100,
         TREE_CONDITION_RATING_PERC > 0,
         DBH_CM > 0) %>%
  ggplot(aes(x = DBH_CM, y = TREE_CONDITION_RATING_PERC)) +
  geom_point(alpha = 0.1) +
  geom_smooth(method = "lm")
```

## Tree Condition vs Size (Diameter of Trunk)
- Coefficient of determination is very low
- Only 0.2% of the data can be described by this linear model
```{r echo=FALSE}
public_trees_predict <- public_trees %>%
  filter(!is.na(TREE_CONDITION_RATING_PERC), !is.na(DBH_CM)) %>%
  filter(TREE_CONDITION_RATING_PERC < 100,
         TREE_CONDITION_RATING_PERC > 0,
         DBH_CM > 0)
predictcondition = lm(TREE_CONDITION_RATING_PERC~DBH_CM, data = public_trees_predict)

# Linear model coefficients
predictcondition$coef

# R squared
rsquared(predictcondition)
```

## Green Ash Proportion
- Determine if there is a difference between the proportions of Green Ash in developing and developed communities
- Does the city know that they should be planting fewer Green Ash?
- Bootstrapping to create confidence intervals

```{r Setup Green Ash Bootstrap, include=FALSE}
# Tally and proportion of trees per community type
name_tally <- public_trees %>%
  filter(!is.na(COMMON_NAME), #Filters NAs from name column
         !is.na(SRG), # Filters NAs from community type column
         SRG != "N/A") %>% # Filters "N/A"s from community type column
  group_by(SRG, COMMON_NAME) %>%
  tally() %>%
  arrange(desc(n)) %>%
  mutate(proportion = n / sum(n))

# Number of Green Ash trees in "Built-Out" Communities
X_GA_built <- pull(name_tally %>% filter(SRG == "BUILT-OUT", COMMON_NAME == "GREEN ASH"), n)

# Number of Green Ash trees in "Developing" Communities
X_GA_developing <- pull(name_tally %>% filter(SRG == "DEVELOPING", COMMON_NAME == "GREEN ASH"), n)

# Total number of trees in "Built-Out" Communities
n_built <- public_trees %>%
  filter(!is.na(COMMON_NAME),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG) %>%
  tally() %>%
  filter(SRG == "BUILT-OUT") %>%
  pull(n)

# Total number of trees in "Developing" Communities
n_developing <- public_trees %>%
  filter(!is.na(COMMON_NAME),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG) %>%
  tally() %>%
  filter(SRG == "DEVELOPING") %>%
  pull(n)
```

## Green Ash Proportion Differences
```{r Green Ash Prop Diff, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for difference in Green Ash proportion
GA_built_vector <- c(rep(0, n_built - X_GA_built), rep(1, X_GA_built))
GA_developing_vector <- c(rep(0, n_developing - X_GA_developing), rep(1, X_GA_developing))
bootstrap_GA_diff <- do(1000) * (mean(resample(GA_developing_vector, n = n_developing)) - mean(resample(GA_built_vector, n = n_built)))
bootstrap_GA_diff_plot <- bootstrap_GA_diff %>% 
  ggplot(aes(x = result)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0001) +
  xlab("Bootstrap Statistic - Difference in Sample Proportions of Green Ash") +
  ylab("Counts") +
  ggtitle("Distribution - Difference in Sample Proportions of Green Ash")
bootstrap_GA_diff_plot
```

## Green Ash Proportion Differences
```{r Green Ash Prop Diff stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~result, data = bootstrap_GA_diff)
mosaic::qdata(~result, c(0.025, 0.975), data = bootstrap_GA_diff)
```

## Green Ash Proportion
- There is a clear difference between the Green Ash proportion in developing and developed communities
- Fewer Green Ash are being planted in developing communities
- Is it because Green Ash condition is on average less than the average overall condition of other trees?

## Picea Proportion
- Determine if there is a difference between the proportions of Picea in developing and developed communities
- Does the city know that they should be planting more Picea?
- Bootstrapping to create confidence intervals
- Fewer Genus categories (fewer naming inconsistencies than common names)

```{r Setup Picea Bootstrap, warning=FALSE, include=FALSE}
# Tally and proportion of trees per community type
genus_tally <- public_trees %>%
  filter(!is.na(GENUS), #Filters NAs from name column
         !is.na(SRG), # Filters NAs from community type column
         SRG != "N/A") %>% # Filters "N/A"s from community type column
  group_by(SRG, GENUS) %>%
  tally() %>%
  arrange(desc(n)) %>%
  mutate(proportion = n / sum(n))

# Proportion of Picea in Developing or Built Out Communities

# Number of Picea trees in "Built-Out" Communities
X_picea_built <- pull(genus_tally %>% filter(SRG == "BUILT-OUT", GENUS == "PICEA"), n)

# Number of Picea trees in "Developing" Communities
X_picea_developing <- pull(genus_tally %>% filter(SRG == "DEVELOPING", GENUS == "PICEA"), n)

# Total number of trees in "Built-Out" Communities
n_built_genus <- public_trees %>%
  filter(!is.na(GENUS),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG) %>%
  tally() %>%
  filter(SRG == "BUILT-OUT") %>%
  pull(n)

# Total number of trees in "Developing" Communities
n_developing_genus <- public_trees %>%
  filter(!is.na(GENUS),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG) %>%
  tally() %>%
  filter(SRG == "DEVELOPING") %>%
  pull(n)
```

## Picea Proportion Differences
```{r Picea Prop Diff, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for difference in Picea proportion
picea_built_vector <- c(rep(0, n_built_genus - X_picea_built), rep(1, X_picea_built))
picea_developing_vector <- c(rep(0, n_developing_genus - X_picea_developing), rep(1, X_picea_developing))
bootstrap_picea_diff <- do(1000) * (mean(resample(picea_developing_vector, n = n_developing_genus)) - mean(resample(picea_built_vector, n = n_built_genus)))
bootstrap_picea_diff_plot <- bootstrap_picea_diff %>% 
  ggplot(aes(x = result)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0005) +
  xlab("Bootstrap Statistic - Difference in Sample Proportions of Picea") +
  ylab("Counts") +
  ggtitle("Distribution - Difference in Sample Proportions of Picea")
bootstrap_picea_diff_plot
```

## Picea Proportion Differences
```{r Picea Prop Diff stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~result, data = bootstrap_picea_diff)
mosaic::qdata(~result, c(0.025, 0.975), data = bootstrap_picea_diff)
```

## Picea Proportion
- There is a clear difference between the Picea proportion in developing and developed communities
- Fewer Picea are being planted in developing communities
- Why is this the case? Shouldn't the city be planting more?
- Are there other reasons for the decline in proportion of Picea?

## Mature Size Proportion
- Determine if there is a difference between the proportions of Small trees in developing and developed communities
- Does the city know that they should be planting more small trees?
- Bootstrapping to create confidence intervals

```{r Setup Size Bootstrap, warning=FALSE, include=FALSE}
# Tally and proportion of trees per community type
size_tally <- public_trees %>%
  filter(!is.na(MATURE_SIZE),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG, MATURE_SIZE) %>%
  tally() %>%
  arrange(desc(n)) %>%
  mutate(proportion = n / sum(n))

# Proportion of Picea in Developing or Built Out Communities

# Number of Small trees in "Built-Out" Communities
X_small_built <- pull(size_tally %>% filter(SRG == "BUILT-OUT", MATURE_SIZE == "SMALL"), n)

# Number of Small trees in "Developing" Communities
X_small_developing <- pull(size_tally %>% filter(SRG == "DEVELOPING", MATURE_SIZE == "SMALL"), n)

# Total number of trees in "Built-Out" Communities
n_built_size <- public_trees %>%
  filter(!is.na(MATURE_SIZE),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG) %>%
  tally() %>%
  filter(SRG == "BUILT-OUT") %>%
  pull(n)

# Total number of trees in "Developing" Communities
n_developing_size <- public_trees %>%
  filter(!is.na(MATURE_SIZE),
         !is.na(SRG),
         SRG != "N/A") %>%
  group_by(SRG) %>%
  tally() %>%
  filter(SRG == "DEVELOPING") %>%
  pull(n)
```

## Small Proportion Differences
```{r Small Prop Diff, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for difference in Picea proportion
small_built_vector <- c(rep(0, n_built_size - X_small_built), rep(1, X_small_built))
small_developing_vector <- c(rep(0, n_developing_size - X_small_developing), rep(1, X_small_developing))
bootstrap_small_diff <- do(1000) * (mean(resample(small_developing_vector, n = n_developing_size)) - mean(resample(small_built_vector, n = n_built_size)))
bootstrap_small_diff_plot <- bootstrap_small_diff %>% 
  ggplot(aes(x = result)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0005) +
  xlab("Bootstrap Statistic - Difference in Sample Proportions of Small Trees") +
  ylab("Counts") +
  ggtitle("Distribution - Difference in Sample Proportions of Small Trees")
bootstrap_small_diff_plot
```

## Small Proportion Differences
```{r Small Prop Diff stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~result, data = bootstrap_small_diff)
mosaic::qdata(~result, c(0.025, 0.975), data = bootstrap_small_diff)
```

## Small Proportion
- There is a clear difference between the proportions of Small trees in developing and developed communities
- More Small trees are being planted in developing communities
- Is it because the condition of Small trees is on average greater than the condition of Large trees?
- We may need to check the proportions of Medium sized trees as well to make an accurate conclusion

## Conclusions
- The City of Calgary Public Trees dataset gives us an understanding of tree condition over a variety of variables
- The most popular tree NAME appear to have a lower average tree condition than the average tree condition of all other trees
- The most popular tree GENUS appear to have a higher average tree condition than the average tree condition of all other trees
- Small trees on average have a higher tree condition than large trees

## Conclusions
- The proportion of Green Ash appears to be reducing in developing communities

    - Good! Based on average tree condition of Green Ash, this is a good trend.
    
- The proportion of Picea appears to be reducing in developing communities

    - Bad! Based on average tree condition of Picea, this is not a good trend.
    

- The proportion of Small trees appears to be increasing in developing communities

    - Good! Based on average tree condition of Small trees, this is a good trend.

## References
- Public Trees

    - City of Calgary (2019) Public Trees [Online]. Available at: https://data.calgary.ca/Environment/Public-Trees/tfs4-3wwa (Accessed: 28 September 2019)
Community Points

    - City of Calgary (2019) Community Points [Online]. Available at: https://data.calgary.ca/Base-Maps/Community-Points/j9ps-fyst (Accessed: 28 September 2019)