---
title: "DATA 603 - Lecture Notes Unit 1 Part 3"
output: html_document
---

```{r Load Packages, include = FALSE}
library(ggplot2)
library(dplyr)
library(readr)
library(olsrr)
library(leaps)
library(GGally)
```



#### Inclass Practice Problem 6

Clerical staff work hours. In any production process in which one or more workers are engaged in a variety of tasks, the total time spent in production varies as a function of the size of the work pool and the level of output of the various activities. For example, in a large metropolitan department store, the number of hours worked (Y) per day by the clerical staff may depend on the following variables:
  - X1 = Number of pieces of mail processed (open, sort, etc.)
  - X2 = Number of money orders and gift certificates sold,
  - X3 = Number of window payments (customer charge accounts) transacted ,
  - X4 = Number of change order transactions processed ,
  - X5 = Number of checks cashed ,
  - X6 = Number of pieces of miscellaneous mail processed on an ‘’as available’’ basis , and
  - X7 = Number of bus tickets sold
The data are provided in CLERICAL.csv file count for these activities on each of 52 working
days. Conduct a Stepwise Regression Procedure and All-Possible-Regressions procedure of
the data using R software package.

```{r Clerical Individual t test}
clerical <- read_csv("/Users/Ellsworth/Documents/School/DATA603/Lectures/CLERICAL.csv")
clerical <- clerical[, -c(1, 2)]
clerical_full <- lm(data = clerical, Y ~.)
summary(clerical_full)
```

Y ~ X2 + X4 + X5

```{r Clerical Stepwise}
stepw_clerical <- ols_step_both_p(clerical_full, pent = 0.1, prem = 0.3, details = TRUE)
```

Y ~ X2 + X4 + X5 + X6

```{r Clerical Best Subset}
# need to install the package leaps for best.subset() function
clerical_best_subset <- regsubsets(Y ~., data = clerical, nv = 10)
summary(clerical_best_subset)
clerical_reg_summary <- summary(clerical_best_subset)
# for the output interpretation
rsquare <- c(clerical_reg_summary$rsq)
cp <- c(clerical_reg_summary$cp)
AdjustedR<-c(clerical_reg_summary$adjr2)
RMSE<-c(clerical_reg_summary$rss)
cbind(rsquare,cp,RMSE,AdjustedR)
```

Y ~ X1 + X2 + X3 + X4 + X5 + X6

```{r Clerical Interaction Individual t test}
clerical_interaction <- lm(Y ~ (X2 + X4 + X5)**2, data = clerical)
summary(clerical_interaction)
```

Y ~ X2 + X4 + X5

```{r Clerical Interaction Stepwise}
clerical_interaction_stepwise <- lm(Y ~ (X2 + X4 + X5 + X6)**2, data = clerical)
summary(clerical_interaction_stepwise)
```

Y ~ X2 + X4 + X5 + X2*X6

```{r Clerical Interaction Best Subset}
clerical_interaction_best_subset <- lm(Y ~ (X1 + X2 + X3 + X4 + X5 + X6)**2, data = clerical)
summary(clerical_interaction_best_subset)
```

Y ~ X1 + X2 + X3 + X4 + X5 + X6 + X1 * X6 + X2 * X6

```{r GGPairs}
#considering high order model between Xs and Y to improve the model
# need to install the GGally package for ggpairs function

#option 1: using function ggpairs()
clericaldata <- data.frame(clerical$Y, clerical$X1, clerical$X2, clerical$X3,
                         clerical$X4, clerical$X5, clerical$X6, clerical$X7)
#ggpairs(salarydata)
#LOESS or LOWESS: LOcally WEighted Scatter-plot Smoother
ggpairs(clericaldata, lower = list(continuous = "smooth_loess", combo =
 "facethist", discrete = "facetbar", na = "na"))
```

Improve individual t test model
```{r}
best_model_t_test <- lm(Y ~ X2 + I(X2^2) + X4 + X5, data = clerical)
summary(best_model_t_test)
best_model_t_test1 <- lm(Y ~ X2 + I(X2^2) + I(X2^3) + X4 + X5, data = clerical)
summary(best_model_t_test1)
# X2^2 is the best
```

Y ~ X2 + X2^2 + X4 + X5

Improve stepwise model
```{r}

```

Y ~ X2 + X2^2 + X4 + X5

Improve best subset model
```{r}

```

Y ~ X1 + X2 + X3 + X4 + X5 + X6 + X1 x X6 + X2 x X6

