---
title: "Additional Project Info"
output: html_notebook
---

## Explore - Tree Species
- Top 10 Tree Species in Calgary

```{r List Top 10 Species, include=FALSE}
# Tally number of trees per Species type
species <- public_trees %>%
  top_n(10) %>%
  filter(!is.na(SPECIES)) %>% # Filters NAs from species column
  group_by(SPECIES) %>%
  tally() %>%
  arrange(desc(n))
species_top10 <- pull(species %>% top_n(10), SPECIES)
```

```{r Top 10 Species, echo=FALSE, warning=FALSE}
# Plot
species_plot <- species %>%
  filter(SPECIES %in% species_top10) %>%
  ggplot(aes(x = reorder(SPECIES, n), y = n / 1000)) +
  geom_col() +
  xlab("Tree Species") +
  ylab("Total Count (Thousands) - Calgary") +
  ggtitle("Tree Count By Species - Top 10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
species_plot
```

## Tree Condition - Top 10 Species
```{r Top 10 Species Conditions, echo=FALSE, warning=FALSE}
species_condition <- public_trees %>%
  filter(!is.na(TREE_CONDITION_RATING_PERC), TREE_CONDITION_RATING_PERC < 100) %>%
  filter(!is.na(SPECIES)) %>%  # Filters NAs from species column
  filter(SPECIES %in% species_top10)
species_boxplot <- species_condition %>%
  ggplot(aes(x = SPECIES, y = TREE_CONDITION_RATING_PERC)) +
  geom_boxplot() +
  xlab("Tree Species") +
  ylab("Tree Condition (%)") +
  ggtitle("Boxplot Tree Condition By Species - Top 10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
species_boxplot
```

## Tree Condition - Top 10 Trees by Name
```{r Top 10 Tree Names Conditions, echo=FALSE, warning=FALSE}
names_condition <- public_trees %>%
  filter(!is.na(TREE_CONDITION_RATING_PERC), TREE_CONDITION_RATING_PERC < 100) %>%
  filter(!is.na(COMMON_NAME)) %>%  # Filters NAs from species column
  filter(COMMON_NAME %in% names_top10)
names_boxplot <- names_condition %>%
  ggplot(aes(x = COMMON_NAME, y = TREE_CONDITION_RATING_PERC)) +
  geom_boxplot() +
  xlab("Tree Names") +
  ylab("Tree Condition (%)") +
  ggtitle("Boxplot Tree Condition By Common Names - Top 10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
names_boxplot
```

## Tree Condition - Top 10 Trees by Genus
```{r Top 10 Genus Conditions, echo=FALSE, warning=FALSE}
genus_condition <- public_trees %>%
  filter(!is.na(TREE_CONDITION_RATING_PERC), TREE_CONDITION_RATING_PERC < 100) %>%
  filter(!is.na(GENUS)) %>%  # Filters NAs from species column
  filter(GENUS %in% genus_top10)
genus_boxplot <- genus_condition %>%
  ggplot(aes(x = GENUS, y = TREE_CONDITION_RATING_PERC)) +
  geom_boxplot() +
  xlab("Tree Genus") +
  ylab("Tree Condition (%)") +
  ggtitle("Boxplot Tree Condition By Genus - Top 10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
genus_boxplot
```

## Green Ash Proportion - Built-out
```{r Green Ash Prop Built, warning=FALSE, include=FALSE}

bootstrap_GA_built <- do(1000) * mean(resample(GA_built_vector, n = n_built))
bootstrap_GA_built_plot <- bootstrap_GA_built %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0001) +
  xlab("Bootstrap Statistic - Sample Proportion of Green Ash in Built-Out Communities") +
  ylab("Counts") +
  ggtitle("Distribution - Sample Proportion of Green Ash - Built-Out Communities")
bootstrap_GA_built_plot
```

## Green Ash Proportion - Built-out
```{r Green Ash Prop Built Stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~mean, data = bootstrap_GA_built)
mosaic::qdata(~mean, c(0.025, 0.975), data = bootstrap_GA_built)
```


## Green Ash Proportion - Developing
```{r Green Ash Prop Developing, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for Green Ash in "Developing" Communities

bootstrap_GA_developing <- do(1000) * mean(resample(GA_developing_vector, n = n_developing))
bootstrap_GA_developing_plot <- bootstrap_GA_developing %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0001) +
  xlab("Bootstrap Statistic - Sample Proportion of Green Ash in Developing Communities") +
  ylab("Counts") +
  ggtitle("Distribution - Sample Proportion of Green Ash - Developing Communities")
bootstrap_GA_developing_plot
```

## Green Ash Proportion - Developing
```{r Green Ash Prop Developing Stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~mean, data = bootstrap_GA_developing)
mosaic::qdata(~mean, c(0.025, 0.975), data = bootstrap_GA_developing)
```

## Picea Proportion - Built-out
```{r Picea Prop Built, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for Picea in "Built-Out" Communities

bootstrap_picea_built <- do(1000) * mean(resample(picea_built_vector, n = n_built_genus))
bootstrap_picea_built_plot <- bootstrap_picea_built %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0001) +
  xlab("Bootstrap Statistic - Sample Proportion of Picea in Built-Out Communities") +
  ylab("Counts") +
  ggtitle("Distribution - Sample Proportion of Picea - Built-Out Communities")
bootstrap_picea_built_plot
```

## Picea Proportion - Built-out
```{r Picea Prop Built stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~mean, data = bootstrap_picea_built)
mosaic::qdata(~mean, c(0.025, 0.975), data = bootstrap_picea_built)
```

## Picea Proportion - Developing
```{r Picea Prop Developing, warning=FALSE, include=FALSE}
# Build bootstrap distribution for Picea in "Developing" Communities

bootstrap_picea_developing <- do(1000) * mean(resample(picea_developing_vector, n = n_developing_genus))
bootstrap_picea_developing_plot <- bootstrap_picea_developing %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0005) +
  xlab("Bootstrap Statistic - Sample Proportion of Picea in Developing Communities") +
  ylab("Counts") +
  ggtitle("Distribution - Sample Proportion of Picea - Developing Communities")
bootstrap_picea_developing_plot 
```

## Picea Proportion - Developing
```{r Picea Prop Developing stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~mean, data = bootstrap_picea_developing)
mosaic::qdata(~mean, c(0.025, 0.975), data = bootstrap_picea_developing)
```

## Small Proportion - Built-out
```{r Small Prop Built, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for Small in "Built-Out" Communities

bootstrap_small_built <- do(1000) * mean(resample(small_built_vector, n = n_built_size))
bootstrap_small_built_plot <- bootstrap_small_built %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0001) +
  xlab("Bootstrap Statistic - Sample Proportion of Small Trees in Built-Out Communities") +
  ylab("Counts") +
  ggtitle("Distribution - Sample Proportion of Small Trees - Built-Out Communities")
bootstrap_small_built_plot
```

## Small Proportion - Built-out
```{r Small Prop Built stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~mean, data = bootstrap_small_built)
mosaic::qdata(~mean, c(0.025, 0.975), data = bootstrap_small_built)
```

## Small Proportion - Developing
```{r Small Prop Developing, echo=FALSE, warning=FALSE}
# Build bootstrap distribution for Small in "Developing" Communities

bootstrap_small_developing <- do(1000) * mean(resample(small_developing_vector, n = n_developing_size))
bootstrap_small_developing_plot <- bootstrap_small_developing %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(col = 'blue', fill = 'red', binwidth = 0.0005) +
  xlab("Bootstrap Statistic - Sample Proportion of Small Trees in Developing Communities") +
  ylab("Counts") +
  ggtitle("Distribution - Sample Proportion of Small Trees - Developing Communities")
bootstrap_small_developing_plot 
```

## Small Proportion - Developing
```{r Small Prop Developing stats, echo=FALSE, warning=FALSE}
# Generate statistics and 95% confidence interval
mosaic::favstats(~mean, data = bootstrap_small_developing)
mosaic::qdata(~mean, c(0.025, 0.975), data = bootstrap_small_developing)
```

## Linear Relationships
- The Public Trees dataset consists of mostly categorical variables
- Three numeric variables to test linear relationships

    - Tree Condition
    - Tree Size (Diameter)
    - Tree Rating

## Tree Condition vs Tree Rating
```{r echo=FALSE}
# Tree Condition vs Tree Rating
public_trees %>%
  filter(!is.na(TREE_CONDITION_RATING_PERC), !is.na(RATING)) %>%
  filter(TREE_CONDITION_RATING_PERC < 100, RATING > 50) %>%
  ggplot(aes(x = RATING, y = TREE_CONDITION_RATING_PERC)) +
  geom_point(alpha = 0.1)
```

## Difference of Tree Condition Means - Green Ash and Other
- Boxplot for a visual comparison
```{r echo=FALSE}
boxplot(GA_conditions$TREE_CONDITION_RATING_PERC, total_conditions_name$TREE_CONDITION_RATING_PERC, names=c("Green Ash", "Other"), ylab="Tree Condition Rating", col='blue', main="Boxplots of Tree Condition Rating (Green Ash Vs. Other)")
```